### 2016-07-04 case-studies.tex

% To understand whether a straightforward implementation of the view
% counter preserves the invariant, we considered a \txnimp program
% that is representative of the counter application. In particular, we
% considered four concurrent transactions that operate on a shared
% counter: a reader transaction that reads the counter twice obtaining
% the values $k_1$ and $k_2$, respectively, and three writer
% transactions, each incrementing the counter once. If the reader
% witnesses a monotonically increasing view count, then $k_2\ge k_1$.
% We then instantiated our reasoning framework for an {\sc sc} store
% with {\sc rc} as the default isolation level (Postgres is one such
% store), with the aim of verifying the program under this
% instantiation.  Our attempts were however unsuccessful owing to our
% inability to prove that a writer transaction always commits a value
% no smaller than the current value of the counter. This points to the
% existence of anamolous executions where  To understand the
% prevalence of such anomalies, we implemented a view counter
% application in Ruby-on-Rails using standard templates and
% techniques.   
To understand the effect of changing the store consistency level, we
reimplemented the counter as a \emph{replicated data type}
(RDT)~\cite{crdt,burckhardt14} on top of Cassandra
\emph{bolted-on}~\cite{bailisbolton} with {\sc cc} and {\sc sc}
consistency levels, and ANSI SQL isolation levels. Keeping the
consistency level at {\sc ec}, we then repeated our experiments with
committed reads and serializable increments - a combination which was
proved to preserve monotonicity on an {\sc sc} store, but nonetheless
observed monotonicity violations. On the other hand, instantiating the
store consistency level with {\sc cc} allowed us to obtain a proof,
and recover the monotonicity invariant. We in fact obtained a proof
after relaxing the serializability requirement for writes, by changing
the definition of the $\interp{\E}(X)$ (\S\ref{sec:opsem}) to return
the value of the largest write-to-$X$ instead of the latest
write-to-$X$.  We implemented this version of the counter RDT, and
obtained a 52\% improvement in latency compared to the previous
implementation\footnote{The tradeoff is that the view count is now no
  longer accurate, although it grows monotonically, thus preserving
  the application invariant.}.

>>>>

% To understand whether a straightforward implementation of the view
% counter preserves the invariant, we considered a \txnimp program
% that is representative of the counter application. In particular, we
% considered four concurrent transactions that operate on a shared
% counter: a reader transaction that reads the counter twice obtaining
% the values $k_1$ and $k_2$, respectively, and three writer
% transactions, each incrementing the counter once. If the reader
% witnesses a monotonically increasing view count, then $k_2\ge k_1$.
% We then instantiated our reasoning framework for an {\sc sc} store
% with {\sc rc} as the default isolation level (Postgres is one such
% store), with the aim of verifying the program under this
% instantiation.  Our attempts were however unsuccessful owing to our
% inability to prove that a writer transaction always commits a value
% no smaller than the current value of the counter. This points to the
% existence of anamolous executions where  To understand the
% prevalence of such anomalies, we implemented a view counter
% application in Ruby-on-Rails using standard templates and
% techniques.   
To understand the effect of changing the store consistency level, we
reimplemented the counter as a \emph{replicated data type}
(RDT)~\cite{crdt,burckhardt14} on top of Cassandra
\emph{bolted-on}~\cite{bailisbolton} with {\sc cc} and {\sc sc}
consistency levels, and ANSI SQL isolation levels. Keeping the
consistency level at {\sc ec}, we then repeated our experiments with
committed reads and serializable increments - a combination which was
proved to preserve monotonicity on an {\sc sc} store, but nonetheless
observed monotonicity violations. On the other hand, instantiating the
store consistency level with {\sc cc} allowed us to obtain a proof, and
recover the monotonicity invariant. We in fact obtained a proof after
relaxing the serializability requirement for writes, by changing the
definition of the $\interp{\E}(X)$ (\S\ref{sec:opsem}) to return the
value of the largest write-to-$X$ instead of the latest write-to-$X$.
We implemented this version of the counter RDT, and obtained a 52\%
improvement in latency compared to the previous
implementation\footnote{The trade-off is that the view count is now no
longer accurate, although it grows monotonically.}.

### 2016-07-04 opsem.tex

Fundamental to our development is the notion of a trace invariant
($\I$). $\I$ is a function from traces ($\E$) to first-order logical
formula ($\texttt{Prop}$) that define well-formedness
constraints over traces.  The machine takes a step only if the
resulting trace satisfies the constraints imposed by $\I$. This
behaviour is captured by the auxiliary reduction rule
\rulelabel{E-Aux} that factors out the trace extension aspect of the
evaluation by abstracting away the operation-specific behaviour as a
function that generates an appropriate effect. We let $\mathcal{F}$
denote this function.  \rulelabel{E-Aux} uses $\mathcal{F}$ to
generate a new effect and extend the trace ($\E = (\A,\visZ)$)
\emph{only if} the well-formedness constraints imposed by $\I$ on $\E$
(i.e., $\I(\E')$) are satisfied. Otherwise, it gets stuck. In an
execution that runs to completion, every small-step preserves the
well-formedness of a trace, thus ensuring the invariance of $\I$.
Note that the semantics makes no assumptions about $\I$ other than its
type. As such, it can be instantiated with any trace-parametric
proposition that expresses constraints over the given trace. For
instance, consider the $\psi_{RC}$ specification from
\S\ref{sec:motivation}, but with bounded $T_1$ and $T_2$ instantiated
with \C{Wd1} and \C{Wd2}, respectively. The instantiated specification
is the following term:

@@

@@

As described previously, \rulelabel{E-Aux} abstracts away the
operation-specific behaviour of a machine step as a function ($\F$)
that accepts a set ($S$) of effects chosen by the machine to make
visible to the operation, interprets the operation w.r.t $S$, and
returns an appropriate effect that encodes its return value. Rules
\rulelabel{E-Var}, \rulelabel{E-Asgn} and \rulelabel{E-Commit} define
such functions for read, write and commit operations, respectively.
The effect returned by the function in each case includes its
transaction id ($T_i$) along with an arbitrarily chosen effect id
($j$) that is later verified to be unique in \rulelabel{E-Aux}. The
$\rval$ for a write is the value being written, and for commit it is
$\bot$. In case of a read, the value read depends on how the read
operation chooses to interpret the given set ($S$) of visible effects.
The interpretation may depend on application semantics. For example, a
monotonically increasing counter application may choose to let a write
with the largest value determine the value of a read. A more
straightforward interpretation has already been described before. To
accommodate multiple interpretations, the semantics is made parametric
over an interpretation function ($\interp{\cdot}$) that accepts a set
of effects and a variable name, and returns the value associated with
the variable. A straightforward interpretation function that chooses
the last write (i.e., write with largest id) is shown below:

>>>>

Fundamental to our development is the notion of a trace invariant
($\I$). $\I$ is a function from traces ($\E$) to first-order logical
formula ($\Prop$) that define well-formedness
constraints over traces.  The machine takes a step only if the
resulting trace satisfies the constraints imposed by $\I$. This
behaviour is captured by the auxiliary reduction rule
\rulelabel{E-Aux} that factors out the trace extension aspect of the
evaluation by abstracting away the operation-specific behaviour as a
function that generates an appropriate effect. We let $\mathcal{F}$
denote this function.  \rulelabel{E-Aux} uses $\mathcal{F}$ to
generate a new effect and extend the trace ($\E = (\A,\visZ)$)
\emph{only if} the well-formedness constraints imposed by $\I$ on $\E$
(i.e., $\I(\E')$) are satisfied. Otherwise, it gets stuck. In an
execution that runs to completion, every small-step preserves the
well-formedness of a trace, thus ensuring the invariance of $\I$.
Note that the semantics makes no assumptions about $\I$ other than its
type. As such, it can be instantiated with any trace-parametric
proposition that expresses constraints over the given trace. For
instance, consider the $\psi_{RC}$ specification from
\S\ref{sec:motivation}, but with bounded $T_1$ and $T_2$ instantiated
with \C{Wd1} and \C{Wd2}, respectively. The instantiated specification
is the following term:

@@

% TRACE RELATIONS
% ---------------
\input{trace-relations}

@@

As described previously, \rulelabel{E-Aux} abstracts away the
operation-specific behaviour of a machine step as a function ($\F$)
that accepts a set ($S$) of effects chosen by the machine to make
visible to the operation, interprets the operation w.r.t. $S$, and
returns an appropriate effect that encodes its return value. Rules
\rulelabel{E-Var}, \rulelabel{E-Asgn} and \rulelabel{E-Commit} define
such functions for read, write and commit operations, respectively.
The effect returned by the function in each case includes its
transaction id ($T_i$) along with an arbitrarily chosen effect id
($j$) that is later verified to be unique in \rulelabel{E-Aux}. The
$\rval$ for a write is the value being written, and for commit it is
$\bot$. In case of a read, the value read depends on how the read
operation chooses to interpret the given set ($S$) of visible effects.
The interpretation may depend on application semantics. For example, a
monotonically increasing counter application may choose to let a write
with the largest value determine the value of a read. To accommodate
multiple interpretations, the semantics is made parametric over an
interpretation function ($\interp{\cdot}$) that accepts a set of
effects and a variable name, and returns the value associated with the
variable. A straightforward interpretation function that chooses the
last write (i.e., write with largest id) is shown below:

### 2016-07-04 reasoning.tex

As mentioned earlier, our proof system is a rely-guarantee framework
that admits compositional reasoning by abstracting away 
environmental interference as a rely relation. A conspicuous
difference between a standard development of rely-guarantee and ours
is that, while the former reasons in terms of program states (variable
to value bindings), we reason in terms of executions as captured by
their traces ($\E$). In particular, our rely ($R$) and guarantee ($G$)
relations relate executions (i.e., $R,\,Q\,\subseteq\,\E\times\E$),
and pre- ($P$) and post- ($Q$) conditions as assertions over executions
(i.e., $P,\,Q\, : \E \rightarrow \Prop$). Our development nonetheless
facilitates state-based reasoning via the interpretation function
($\interp{\cdot}$) introduced in \S\ref{sec:syntax}. For example, if a
\emph{bi-state} rely relation relates every pair of states $\sigma$
and $\sigma'$ such that $\sigma'(X) \ge \sigma(X) \ge 0$, the
corresponding \emph{bi-execution} rely relation relates every pair of
executions $\E$ and $\E'$ such that $\interp{\E}(X) \ge
\interp{\E'}(X) \ge 0$.  Assertions on states are also written
similarly. For instance, consider the post-condition
$\C{C\,=\,k-a1-a2}$ of the program in Fig.~\ref{fig:motiv-eg-1}. The
corresponding assertion on the post-state
($\lambda\sigma.~\sigma(\C{C}) \,=\, \C{k-a1-a2}$) asserts that in all
states resulting from executing the program, the value of \C{C} is
\C{k-a1-a2}. The equivalent execution-based assertion
($\lambda\E.~\interp{\E}(\C{C})\,=\,\C{k-a1-a2}$) asserts that in all
executions of the program, the value of \C{C} is \C{k-a1-a2}.
However, having access to an execution facilitates assertions that go
beyond the state. An example is an invariant of \C{Wd1} described
abstractly in \S\ref{sec:motivation}, and reified as an
execution-based assertion below:

@@

The standard rely-guarantee judgment is a quintuple
$\rg{P,R}{c}{G,Q}$, which informally asserts that if a
command $c$ is executed in a state that satisfies its pre-condition
$P$, provided that every interference step during the execution is
contained inside the rely relation $R$, then the effect that each step
of executing $c$ has on the state is captured by $G$, and the final
state of execution satisfies the post-condition $Q$. The terms
``execution'' and ``step'' are defined with respect to an operational
semantics, which, in our case, is parameterized on the trace invariant
$\I$. Consequently, our rely-guarantee judgment for commands is the
sextuple $\rg{\I,P,R}{c}{G,Q}$, whose semantics differs from the
quintuple in that it requires every step of the execution to additionally preserve
the trace invariant $\I$ for the post-condition to be valid.  As usual,
if $c$ is inside a transaction $T_i$, we write $\txnbox{c}_i$. \txnimp
expressions are side-effecting (they generate \C{RD} effects), and
admit interference during their evaluation.  Therefore, like 
commands, expressions need to be reasoned about explicitly.  In contrast to
commands, however, expressions evaluate to values, and the reasoning
framework should also admit assertions on such values.  We therefore
define a separate judgment for expressions - the septuple
$\rg{\I,P,R}{e}{G,C,Q}$. The new entrant $C : \mathbb{N} \rightarrow
\Prop$ is an assertion on the return value of $e$.

>>>>

As mentioned earlier, our proof system is a rely-guarantee framework
that admits compositional reasoning by abstracting away environmental
interference as a rely relation. A conspicuous difference between a
standard development of rely-guarantee and ours is that, while the
former reasons in terms of program states (variable to value
bindings), we reason in terms of executions as captured by their
traces ($\E$). In particular, our rely ($R$) and guarantee ($G$)
relations relate executions (i.e., $R,\,Q\,\subseteq\,\E\times\E$),
and our pre- ($P$) and post- ($Q$) conditions are assertions over
executions (i.e., $P,\,Q\, : \E \rightarrow \Prop$). Our development
also facilitates state-based reasoning via the interpretation function
($\interp{\cdot}$) introduced in \S\ref{sec:syntax}. For example, if a
\emph{bi-state} rely relation relates every pair of states $\sigma$
and $\sigma'$ such that $\sigma'(X) \ge \sigma(X) \ge 0$, the
corresponding \emph{bi-execution} rely relation relates every pair of
executions $\E$ and $\E'$ such that $\interp{\E}(X) \ge
\interp{\E'}(X) \ge 0$.  Assertions on states are also written
similarly. For instance, consider the post-condition
$\C{C\,=\,k-a1-a2}$ of the program in Fig.~\ref{fig:motiv-eg-1}. The
corresponding assertion on the post-state
($\lambda\sigma.~\sigma(\C{C}) \,=\, \C{k-a1-a2}$) asserts that in all
states resulting from executing the program, the value of \C{C} is
\C{k-a1-a2}. The equivalent execution-based assertion
($\lambda\E.~\interp{\E}(\C{C})\,=\,\C{k-a1-a2}$) asserts that in all
executions of the program, the value of \C{C} is \C{k-a1-a2}.
However, having access to an execution facilitates assertions that go
beyond the state. An example is an invariant of \C{Wd1} described
abstractly in \S\ref{sec:motivation}, and reified as an
execution-based assertion below:

@@

The standard rely-guarantee judgment is a quintuple
$\rg{P,R}{c}{G,Q}$, which informally asserts that if a command $c$ is
executed in a state that satisfies its pre-condition $P$, provided
that every interference step during the execution is contained inside
the rely relation $R$, then the effect that each step of executing $c$
has on the state is captured by $G$, and the final state of execution
satisfies the post-condition $Q$. The terms ``execution'' and ``step''
are defined with respect to an operational semantics, which, in our
case, is parameterized on the trace invariant $\I$. Consequently, our
rely-guarantee judgment for commands is the sextuple
$\rg{\I,P,R}{c}{G,Q}$, whose semantics differs from the quintuple in
that it requires every step of the execution to additionally preserve
the trace invariant $\I$ for the post-condition to be valid.  As
usual, if $c$ is inside a transaction $T_i$, we write $\txnbox{c}_i$.
\txnimp expressions are side-effecting (they generate \C{RD} effects),
and admit interference during their evaluation.  Therefore, like
commands, expressions need to be reasoned about explicitly. But,
unlike commands, expressions evaluate to values, and the reasoning
framework should also admit assertions on such values.  We therefore
define a separate judgment for expressions - the septuple
$\rg{\I,P,R}{e}{G,C,Q}$. The new entrant $C : \mathbb{N} \rightarrow
\Prop$ is an assertion on the return value of $e$.

### 2016-07-04 related-work.tex

\paragraph{Specifying weak isolation.}
Adya~\cite{adyaphd} specifies several weak isolation levels in terms
of \emph{dependency graphs} between transactions, by defining the
kinds of dependency cycles that are forbidden in each case. The
operational nature of Adya's specifications make them suitable for
runtime monitoring and anamoly
detection~\cite{kemmevldb,feketesigmod08,pssi2011}, whereas the
declarative nature of our specifications make them suitable for formal
reasoning about program behaviour. Sivaramakrishnan \emph{et
  al.}~\cite{pldi15} specify isolation levels declaratively as trace
well-formedness conditions, but their specifications implicitly assume
a complete trace with only committed transactions, and is thus not
suitable to reason operationally about a program as it builds the
trace; their framework also requires application invariants be
expressed in terms of the visibility relations used to define
isolation and consistency properties.  Cerone \emph{et
  al.}~\cite{gotsmanconcur15} present a framework to specify isolation
levels with atomic visibility, but their specifications are also for
complete traces; moreover, their formulation implicitly assumes
homogenity, \emph{i.e.}, they require all transactions to execute
using either {\sc si} or {\sc ser}, which is often not the case in
practice. While all the aforementioned specification frameworks use
the vocabulary introduced in~\cite{burckhardt14} to specify replicated
data types and different forms of weak consistency, none of them come
with an associated reasoning framework that can use such
specifications productively towards verifying programs under weak
isolation.

>>>>

\paragraph{Specifying weak isolation.}
Adya~\cite{adyaphd} specifies several weak isolation levels in terms
of \emph{dependency graphs} between transactions, by defining the
kinds of dependency cycles that are forbidden in each case. The
operational nature of Adya's specifications make them suitable for
runtime monitoring and anomaly
detection~\cite{kemmevldb,feketesigmod08,pssi2011}, whereas the
declarative nature of our specifications make them suitable for formal
reasoning about program behaviour. Sivaramakrishnan \emph{et
  al.}~\cite{pldi15} specify isolation levels declaratively as trace
well-formedness conditions, but their specifications implicitly assume
a complete trace with only committed transactions, and is thus not
suitable to reason operationally about a program as it builds the
trace; their framework also requires application invariants be
expressed in terms of the visibility relations used to define
isolation and consistency properties.  Cerone \emph{et
  al.}~\cite{gotsmanconcur15} present a framework to specify isolation
levels with atomic visibility, but their specifications are also for
complete traces; moreover, their formulation implicitly assumes
homogenity, \emph{i.e.}, they require all transactions to execute
using either {\sc si} or {\sc ser}, which is often not the case in
practice. While all the aforementioned specification frameworks use
the vocabulary introduced in~\cite{burckhardt14} to specify replicated
data types and different forms of weak consistency, none of them come
with an associated reasoning framework that can use such
specifications productively towards verifying programs under weak
isolation.

### 2016-07-04 store-consistency.tex

Notably, our view of replication does not explicitly involve replicas.
Fig.~\ref{fig:ec-theirs-vs-ours} contrasts our model of
weakly-consistent replication with a conventional replica-based model.
Under our model, the notion of a replica is subsumed by the concept of
visibility; a replica is defined by the subset ($S$) of global state
($\E.\A$) that an operation witnesses. Constraints over replica states
therefore manifest as constraints over a specific visibility relation.
For example, instead of requiring the store to be causally consistent,
an operation can witness a causally consistent subset of the state;
such demands can be made via the trace invariant $\I$. For a
pre-condition ($P$) of the operation to be useful, it has to be an
assertion over every causally consistent subset of the global state.
Since any replica that eventually executes the operation has to expose
one such subset ($S$), the pre-condition is guaranteed to hold
regardless of the replica. There is however one problem with this
explanation - by considering subsets of just one global state, it
ignores the fact that the global state (hence, the replica states)
change during the execution of the operation. To account for such
changes, we might choose to distinguish between effect generation
event at one replica $r_1$ and effect merge event at replica $r_2$,
requiring that \emph{non-conflicting} operations execute between these
two events at $r_2$, and that they preserve certain
invariants~\cite{gotsmanpopl16}.  Instead, our framework folds all such
machinery into a stability condition predicated on $\I$
(\S\ref{sec:rely-guarantee}).  Since any change to the global state
during the execution of the operation is an interference, and $P$ is
required to be stable with respect to any such interference, it
follows that $P$ is valid on every replica, thus ensuring that
assumptions made at a generation event is also valid at the merge
event.

>>>>

Notably, our view of replication does not explicitly involve replicas.
Fig.~\ref{fig:ec-theirs-vs-ours} contrasts our model of
weakly-consistent replication with a conventional replica-based model.
Under our model, the notion of a replica is subsumed by the concept of
visibility; a replica is defined by the subset ($S$) of global state
($\E.\A$) that an operation witnesses. Constraints over replica states
therefore manifest as constraints over a specific visibility relation.
For example, instead of requiring the store to be causally consistent,
an operation can witness a causally consistent subset of the state;
such demands can be made via the trace invariant $\I$. For a
pre-condition ($P$) of the operation to be useful, it has to be an
assertion over every causally consistent subset of the global state.
Since any replica that eventually executes the operation has to expose
one such subset ($S$), the pre-condition is guaranteed to hold
regardless of the replica. There is however one problem with this
explanation - by considering subsets of just one global state, it
ignores the fact that the global state (hence, the replica states)
change during the execution of the operation. To account for such
changes, we might choose to distinguish between an effect generation
event at one replica $r_1$ and an effect merge event at replica $r_2$,
requiring that \emph{non-conflicting} operations execute between these
two events at $r_2$, and that they preserve certain
invariants~\cite{gotsmanpopl16}.  Instead, our framework folds all such
machinery into a stability condition predicated on $\I$
(\S\ref{sec:rely-guarantee}).  Since any change to the global state
during the execution of the operation is an interference, and $P$ is
required to be stable with respect to any such interference, it
follows that $P$ is valid on every replica, thus ensuring that
assumptions made at a generation event is also valid at the merge
event.
